---
title: "Survival of the Fittest: Cox Regression Analysis"
author: "Kayla Boyd"
format: html
editor: visual
theme: quartz
bibliography: references.bib
csl: ieee.csl
link-external-icon: true
number-sections: true
code-overflow: wrap
code-fold: true
code-summary: "Code"
indent: true
---

[Slides](https://kayla-j-boyd.github.io/IDC6940_hazardousconditions/HazardousCond_slides.html#/title-slide)

# Introduction

The Cox Proportional Hazards Model, often referred to as Cox Regression, is a widely used statistical method for analyzing and modeling time-to-event data. Developed by Sir David Cox in 1972, this semi-parametric model is a powerful tool for understanding the relationship between predictors and the time until a specified event occurs, such as turnover, equipment failure, or customer churn.

## Key Characteristics

-   Semi-parametric Nature: Cox Regression does not assume a specific probability distribution for survival times. Instead, it focuses on the hazard ratio to evaluate the effect of predictors [@Zhang2016].

-   Time-to-Event Analysis: It models the risk (hazard) of an event occurring at a specific time, considering that the event has not occurred before that time.

-   Handles Censoring: Censoring occurs when the event hasn’t occurred for some subjects by the end of the study. Cox Regression effectively accounts for this.

## Applications of Cox Regression

Cox Regression is used across various fields to analyze survival data and event risks:

### Healthcare and Epidemiology

-   Studying survival rates in patients (e.g., cancer or heart disease).

-   Evaluating the effect of treatments, such as comparing drug efficacy [@Hosmer2008].

### Business and Economics

-   Predicting customer churn or employee turnover.

-   Analyzing time to equipment failure in operational systems.

### Engineering and Reliability

-   Modeling the lifespan of mechanical systems or components.

-   Assessing risks in failure times of industrial systems.

### Social Sciences

-   Examining time until marriage, job changes, or retirement.

-   Investigating relapse rates in criminology studies.

## Limitations of Cox Regression

While Cox Regression is versatile, it has limitations:

1.  **Proportional Hazards Assumption**:

    -   The model assumes that the hazard ratios between groups are constant over time. Violations of this assumption can lead to incorrect conclusions [@Hosmer2008].

    -   *Example*: The effect of a predictor (like a treatment) may decrease or increase over time.

2.  **Inability to Model Complex Time-Dependent Effects**:

    -   The standard Cox model struggles with internal time-varying covariates (e.g., changing patient conditions) without additional adjustments.

3.  **Difficulty with Small Sample Sizes**:

    -   Requires adequate event numbers for reliable estimates.

4.  **No Prediction of Exact Survival Times**:

    -   It estimates relative risks and hazard ratios but does not predict when an event will occur.

5.  **Overfitting**:

    -   Including too many covariates relative to the number of events can lead to overfitting.

# Methods

In this study, we utilize the Cox Proportional Hazards Model to examine factors influencing survival. Cox Regression is a powerful tool in survival analysis that models the relationship between the survival time of subjects and one or more predictor variables. Survival can refer to development of a symptom, time to relapse, or time to death [@Walters2012]. Its appeal lies in its semi-parametric nature, as it does not assume a specific baseline hazard function, making it robust and popular in diverse fields like medicine, engineering, and social sciences.

## Model Assumptions

A key assumption of the Cox model is the proportional hazards assumption, which asserts that the effects of different variables on the outcome (survival) are constant over time. This means that, although the baseline hazard function can vary over time, the effect of covariates on the hazard is multiplicative and time-independent. Another essential assumption pertains to censoring, where it is assumed that censoring is non-informative, meaning it is unlikely that the event of interest will be observed in all patients and missing or erroneous results can therefore not be included. This differs from logistic or linear regression.

## Model Variants

The primary model employed in this study is the standard Cox Proportional Hazards model, which estimates the hazard function, $h(t | X) = h_0(t) \exp(\beta X)$, where $h_0(t)$represents the baseline hazard and $\beta$ represents the regression coefficients. In cases where the proportional hazards assumption does not hold, we may consider extensions like stratified Cox models, which allow for covariates that stratify the baseline hazard, or time-dependent Cox models, which can handle time-varying covariates.

## Testing Proportional Hazards

To verify the proportional hazards assumption, we employ the Schoenfeld residuals test, a common diagnostic that assesses correlation between residuals and time. Graphical methods, such as log-minus-log survival plots, are also used to visually inspect proportional hazards. If these diagnostics suggest a violation of the proportional hazards assumption, adjustments are made, including consideration of time-varying covariates.

The image below shows an example of the graphical method using a Kaplan-Meier Survival Plot.

```{r, warning=FALSE, message=FALSE}
# Load necessary libraries
library(survival)
library(survminer)

# Load the lung dataset
data(lung)

# Fit a Kaplan-Meier survival curve
fit <- survfit(Surv(time, status) ~ sex, data = lung)

# Plot the Kaplan-Meier survival curves
ggsurvplot(fit, data = lung, 
           pval = TRUE, 
           conf.int = TRUE, 
           risk.table = TRUE, 
           ggtheme = theme_minimal(), 
           legend.title = "Sex",
           legend.labs = c("Male", "Female"))


```

## Interpretation of Outcomes

The output of a Cox model is typically interpreted in terms of hazard ratios. For each predictor, the hazard ratio describes the change in the hazard of the event (e.g., death, failure) for a one-unit increase in the predictor, holding all other factors constant. A hazard ratio greater than one indicates an increased risk, while a ratio less than one suggests a protective effect. If the risk factor is binary, the result can be interpreted as the estimated increase in the hazard ratio in patients with the risk factor vs. those without [@Hafeez2021]. P-values and confidence intervals around hazard ratios help determine the statistical significance and reliability of these effects.

## Addressing Time-Varying Coefficients

To account for non-proportionality when it is detected, time-varying covariates may be introduced [@Zhang2018]. This involves creating interaction terms between covariates and functions of time or using an extended Cox model framework. This approach allows the impact of certain variables on the hazard to change over time, enhancing the model's flexibility and accuracy when the proportional hazards assumption does not strictly hold.

# Data Analysis and Results

## Dataset Description

We will be applying our method to this dataset from [Kaggle](https://www.kaggle.com/datasets/davinwijaya/employee-turnover/data) that contains real data about employees at a Russian company. There are 16 variables in the dataset, both discrete and continuous, shown in the description table below.

| Variable     | Description                                      |
|--------------|--------------------------------------------------|
| stag         | Experience (time)                                |
| event        | Employee turnover                                |
| gender       | Employee's gender                                |
| age          | Employee's age (year)                            |
| industry     | Employee's industry                              |
| profession   | Employee's profession                            |
| traffic      | From what pipelene employee came to the company. |
| coach        | Presence of a coach (training) on probation      |
| head_gender  | Head (supervisor) gender                         |
| greywage     | The salary does not seem to the tax authorities. |
| way          | Employee's way of transportation                 |
| extraversion | Extroversion score                               |
| independ     | Independence score                               |
| selfcontrol  | Self control score                               |
| anxiety      | Anxiety score                                    |
| novator      | Innovator score                                  |

## Data Preparation

Read the data from our .csv file, load it, and take a quick look at some of the data.

```{r, warning=FALSE, message=FALSE}

library(readr)
turnover <- read_csv("turnover.csv")
head(turnover)

library(cowplot)
library(broom)
library(tidyverse) 
library(tidyr)
library(dplyr)
library(corrplot)
library(RColorBrewer)
library(ggfortify)
library(riskRegression)
library(survival)
library(stringr)
library(zoo)
library(ranger)
library(ggplot2)
library(readxl)
library(MASS)
library(ADGofTest)
library(survminer)
library(car)
library(gtsummary)
library(ggplot2)
library(gridExtra)
library(janitor)
library(magrittr)
library(gt)
library(knitr)
library(kableExtra)
```

Let's check for missing and null values, censored observations (where the event being studied did not occur), and check the distribution of the data.

```{r, warning= FALSE, message=FALSE}
# Check for null and duplicate values

cat("Number of missing values :", sum(is.na(turnover)))
cat("Number of duplicates: ", turnover %>%
    duplicated() %>%
    sum())
# Remove duplicate values

turnover <- unique(turnover)

# Check distribution of data

turnover %>%
  ggplot(aes(x = stag, color = factor(event),
                              fill = factor(event))) +
  geom_histogram(aes(y = ..density..), alpha = 0.5) +
  geom_density(alpha = 0.05) + scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") + theme_minimal() + 
  theme(legend.position = "top")

# Check for censored observations

n <- dim(turnover)[1]
cat((n - sum(turnover$event))/n * 100, "% of observations are censored")


```

We can see that our data is skewed heavily to the right, we have no null values, and 49.8% of employees were not terminated.

Now we will take a look at our covariates and separate them based on whether they are discrete or continuous in order to better visualize the distributions.

```{r, message=FALSE}
# Separate variable based on type

NUM_COLS <- c("stag", "age", "extraversion", "independ", "selfcontrol", "anxiety", "novator")
CAT_COLS <- c("gender", "industry", "profession", "traffic", "coach", "head_gender", "greywage", "way")

# Transform CAT_COLS into categorical type

    for (COL in CAT_COLS){
    turnover[COL] <- turnover[COL] %>% unlist() %>% factor()}
```

### Categorical Covariates

```{r, message=FALSE}
# Categorical covariates

# Selection of covariates (that are only discrete)
turnover.cat <- turnover %>% 
                select_if(is.factor) %>% 
                mutate(event = turnover$event)

turnover.cat %>% ggplot(aes(x = gender, color = factor(event),
                            fill = factor(event))) +
  geom_bar(alpha = 0.5) +
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top") + theme(panel.spacing.x = unit(30, "mm"))
turnover.cat %>% ggplot(aes(x = industry, color = factor(event),
                            fill = factor(event))) +
  geom_bar(alpha = 0.5) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top") + theme(panel.spacing.x = unit(30, "mm"))
turnover.cat %>% ggplot(aes(x = profession, color = factor(event),
                            fill = factor(event))) +
  geom_bar(alpha = 0.5) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top") + theme(panel.spacing.x = unit(30, "mm"))
turnover.cat %>% ggplot(aes(x = traffic, color = factor(event),
                            fill = factor(event))) +
  geom_bar(alpha = 0.5) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top") + theme(panel.spacing.x = unit(30, "mm"))
turnover.cat %>% ggplot(aes(x = coach, color = factor(event),
                            fill = factor(event))) +
  geom_bar(alpha = 0.5) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top") + theme(panel.spacing.x = unit(30, "mm"))
turnover.cat %>% ggplot(aes(x = head_gender, color = factor(event),
                            fill = factor(event))) +
  geom_bar(alpha = 0.5) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top") + theme(panel.spacing.x = unit(30, "mm"))
turnover.cat %>% ggplot(aes(x = greywage, color = factor(event),
                            fill = factor(event))) +
  geom_bar(alpha = 0.5) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top") + theme(panel.spacing.x = unit(30, "mm"))
turnover.cat %>% ggplot(aes(x = way, color = factor(event), 
                                     fill = factor(event))) +
  geom_bar(alpha = 0.5) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top") + theme(panel.spacing.x = unit(30, "mm"))


```

### Continuous Covariates

```{r, message=FALSE}
# Continuous covariates
turnover %>% ggplot(aes(x = stag, color = factor(event), 
                        fill = factor(event))) +
  geom_histogram(aes(y = ..density..), alpha = 0.5) + 
  geom_density(alpha = 0.05) +  
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top")
turnover %>% ggplot(aes(x = event, color = factor(event),
                        fill = factor(event))) +
  geom_histogram(aes(y = ..density..), alpha = 0.5) + 
  geom_density(alpha = 0.05) +
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") + 
  theme_minimal() + theme(legend.position = "top")
turnover %>% ggplot(aes(x = age, color = factor(event),
                        fill = factor(event))) +
  geom_histogram(aes(y = ..density..), alpha = 0.5) +
  geom_density(alpha = 0.05) +
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") + 
  theme_minimal() + theme(legend.position = "top")
turnover %>% ggplot(aes(x = extraversion, color = factor(event), 
                        fill = factor(event))) + 
  geom_histogram(aes(y = ..density..), alpha = 0.5) + 
  geom_density(alpha = 0.05) + scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") + theme_minimal() + 
  theme(legend.position = "top")
turnover %>% ggplot(aes(x = independ, color = factor(event), 
                        fill = factor(event))) +
  geom_histogram(aes(y = ..density..), alpha = 0.5) + 
  geom_density(alpha = 0.05) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") + 
  theme_minimal() + theme(legend.position = "top")
turnover %>% ggplot(aes(x = selfcontrol, color = factor(event),
                        fill = factor(event))) +
  geom_histogram(aes(y = ..density..), alpha = 0.5) +
  geom_density(alpha = 0.05) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top")
turnover %>% ggplot(aes(x = anxiety, color = factor(event),
                        fill = factor(event))) +
  geom_histogram(aes(y = ..density..), alpha = 0.5) +
  geom_density(alpha = 0.05) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top")
turnover %>% ggplot(aes(x = novator, color = factor(event),
                        fill = factor(event))) +
  geom_histogram(aes(y = ..density..), alpha = 0.5) +
  geom_density(alpha = 0.05) + 
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  theme_minimal() + theme(legend.position = "top")
```

From the visualizations, we can see a few variables that might contribute to employee turnover based on their distributions. Below are the observations:

-   Gender, profession, and way seem to be significant factor given the structure of distribution.

-   We can see our event (termination) occurs almost evenly throughout our dataset.

-   Age and stag are heavily right-skewed, most employees being under 40 and tenure being less than 5 years.

Before we move on we need to plot the Kaplan-Meier Survival Plots to ensure we meet the assumption of proportional hazards.

### Proportional Hazards Assumption

#### Gender

```{r}
km_gender<-survfit(Surv(stag, event)~gender, data = turnover, type="kaplan-meier")

ggsurvplot(km_gender, data=turnover,
           conf.int = FALSE,
           ggtheme = theme_minimal(),
           legend.labs = c("female", "male"),
           pval = TRUE,
           pval.method = TRUE)+
  ggtitle("Survival curve based on Gender")
```

#### Industry

```{r}
km_industry<-survfit(Surv(stag, event)~industry, data = turnover, type ="kaplan-meier")
ggsurvplot(km_industry, data=turnover,
           conf.int = FALSE,
           ggtheme = theme_minimal(),
           pval = TRUE,
           pval.method = TRUE)+
  ggtitle("Survival curve based on Industry")

```

#### Profession

```{r}
km_profession<-survfit(Surv(stag, event)~profession, data = turnover, type="kaplan-meier")
ggsurvplot(km_profession, data=turnover,
           conf.int = FALSE,
           ggtheme = theme_minimal(),
           pval = TRUE,
           pval.method = TRUE)+
  ggtitle("Survival curve based on Profession")
```

#### Traffic

```{r}
km_traffic<-survfit(Surv(stag, event)~traffic, data= turnover, type="kaplan-meier")
ggsurvplot(km_traffic, data=turnover,
           conf.int = FALSE,
           ggtheme = theme_minimal(),
           pval = TRUE,
           pval.method = TRUE)+
  ggtitle("Survival curve based on Traffic")
```

#### Coach

```{r}
km_coach<-survfit(Surv(stag, event)~coach, data = turnover, type="kaplan-meier")
ggsurvplot(km_coach, data=turnover,
           conf.int = FALSE,
           ggtheme = theme_minimal(),
           pval = TRUE,
           pval.method = TRUE)+
  ggtitle("Survival curve based on Coach")
```

#### Head_gender

```{r}
km_headgender<-survfit(Surv(stag, event)~head_gender, data= turnover, type="kaplan-meier")
ggsurvplot(km_headgender, data=turnover,
           conf.int = FALSE,
           ggtheme = theme_minimal(),
           legend.labs = c("female", "male"),
           pval = TRUE,
           pval.method = TRUE)+
  ggtitle("Survival curve based on Head Gender")
```

#### Greywage

```{r}
km_greywage<-survfit(Surv(stag, event)~greywage, data = turnover, type ="kaplan-meier")
ggsurvplot(km_greywage, data=turnover,
           conf.int = FALSE,
           ggtheme = theme_minimal(),
           legend.labs = c("grey", "white"),
           pval = TRUE,
           pval.method = TRUE)+
  ggtitle("Survival curve based on Greywage")
```

#### Way

```{r}
km_way<-survfit(Surv(stag, event)~way,
                data= turnover,
                type="kaplan-meier")
ggsurvplot(km_way, data=turnover,
           conf.int = FALSE,
           ggtheme = theme_minimal(),
           legend.labs=c("bus", "car", "foot"),
           pval = TRUE,
           pval.method = TRUE)+
  ggtitle("Survival curve based on Commuters(way)")
```

Based on the above survival curve plots, we conclude that the *gender, coach*, and *head_gender* variables do not meet the assumption of proportional hazards.

## Statistical Modeling

### Crude Model

```{r}
model0<-coxph(Surv(stag, event)~.,
                 data = turnover)
summary(model0)
```

### Refining the Model

We want to refine the model to use the independent variable(s) with the greatest influence from our crude model. To do this, we will use the "step" function in R which iteratively adds or removes variables based on a chosen criterion. For this model we will use the AIC value. The Akaike Information Criterion (AIC) is a statistical metric that evaluates how well a model fits a given set of data.

```{r}
step(model0)
```

Using these results, we can see that the model including independent variables *age, industry, profession, traffic, greywage, selfcontrol*, and *anxiety* produced the most accurate survival model. We will use these variables in our adjusted model.

### Adjusted Model

```{r}
model1<-coxph(Surv(stag, event)~age + industry + profession + traffic + greywage + way + selfcontrol + anxiety,
                data = turnover)
summary(model1)
```

### Hazard Ratios

Hazard ratios above 1 indicate a higher likelihood of the individuals in that category to experience the event (termination) in the given timeframe.

```{r}
hr=exp(model1$coefficients)
hr
```

Our results show that age increases the likelihood of employee turnover, as well as certain professions such as Commercial, Engineering, and PR.

## Results

|   | Crude Model | Adjusted Model |
|-----------------------|:----------------------:|:----------------------:|
| Covariate | Hazard ratio (95% CI), p-value | Hazard ratio (95% CI), p-value |
| genderm | (0.70, 1.15) 0.384 | Excluded |
| **age** | **(1.01, 1.04) 0.002** | **(1.01, 1.03) 0.001** |
| industryBanks | (0.39, 1.64) 0.537 | (0.37, 1.52) 0.424 |
| industryBuilding | (0.37, 1.73) 0.562 | (0.36, 1.64) 0.494 |
| industryConsult | (0.32, 1.42) 0.304 | (0.31, 1.31) 0.215 |
| industryetc | (0.27, 1.18) 0.128 | (0.25, 1.08) 0.080 |
| industryHoReCa | (0.18, 1.53) 0.239 | (0.16, 1.32) 0.149 |
| **industryIT** | **(0.14, 0.66) 0.002** | **(0.13, 0.61) 0.001** |
| **industrymanufacture** | **(0.22, 0.94) 0.032** | **(0.20, 0.85) 0.016** |
| industryMining | (0.23, 1.32) 0.179 | (0.21,1.23) 0.136 |
| **industryPharma** | **(0.14, 0.94) 0.036** | **(0.14, 0.89) 0.027** |
| **industryPowerGeneration** | **(0.15, 0.86) 0.022** | **(0.14, 0.80) 0.013** |
| **industryRealEstate** | **(0.06, 0.56) 0.003** | **(0.05, 0.51) 0.001** |
| **industryRetail** | **(0.17, 0.72) 0.004** | **(0.16, 0.66) 0.002** |
| industryState | (0.23,1.15) 0.104 | (0.22, 1.05) 0.066 |
| **industryTelecom** | **(0.13, 0.73) 0.008** | **(0.12, 0.67) 0.004** |
| **industrytransport** | **(0.18, 0.99) 0.046** | **(0.18, 0.96) 0.041** |
| professionBusinessDevelopment | (0.67, 4.94) 0.237 | (0.68, 4.83) 0.238 |
| **professionCommercial** | **(1.00, 7.34) 0.049** | **(1.03, 7.27) 0.043** |
| professionConsult | (0.64, 4.91) 0.272 | (0.64, 4.70) 0.281 |
| professionEngineer | (0.94, 7.80) 0.063 | (0.92, 7.22) 0.072 |
| professionetc | (0.63, 4.22) 0.317 | (0.61, 4.07) 0.343 |
| professionFinance | (0.377, 2.96) 0.917 | (0.38, 2.92) 0.910 |
| professionHR | (0.53, 2.84) 0.638 | (0.54, 2.87) 0.603 |
| professionIT | (0.41, 2.81) 0.887 | (0.41, 2.61) 0.951 |
| professionLaw | (0.42, 5.33) 0.533 | (0.39, 4.82) 0.621 |
| **professionmanage** | **(1.36, 9.63) 0.010** | **(1.36, 9.57) 0.01** |
| professionMarketing | (0.80, 5.32) 0.132 | (0.79, 5.16) 0.141 |
| professionPR | (0.66, 8.17) 0.186 | (0.65, 7.95) 0.197 |
| professionSales | (0.66, 4.15) 0.280 | (0.68, 4.06) 0.266 |
| professionTeaching | (0.61, 5.66) 0.278 | (0.61, 5.70) 0.270 |
| **trafficempjs** | **(1.37, 4.69) 0.003** | **(1.28, 4.32) 0.006** |
| trafficfriends | (0.58, 2.21) 0.721 | (0.54, 2.00) 0.917 |
| trafficKA | (0.58, 2.30) 0.688 | (0.56, 2.20) 0.769 |
| trafficrabrecNErab | (0.94, 3.18) 0.077 | (0.89, 2.95) 0.115 |
| trafficrecNErab | (0.45, 2.01) 0.893 | (0.42, 1.85) 0.736 |
| trafficreferal | (0.76, 2.73) 0.257 | (0.72, 2.55) 0.337 |
| **trafficyoujs** | **(1.05, 3.53) 0.034** | **(1.00, 3.33) 0.049** |
| coachno | (0.85,1.316) 0.615 | Excluded |
| coachyes | (0.92, 1.66) 0.164 | Excluded |
| head_genderm | (0.86, 1.29) 0.590 | Excluded |
| **greywagewhite** | **(0.46, 0.79) 0.0002** | **(0.46, 0.78) 0.0001** |
| **waycar** | **(0.67, 1.00) 0.052** | **(0.66, 0.99) 0.039** |
| **wayfoot** | **(0.48, 0.94) 0.020** | **(0.49, 0.96) 0.029** |
| extraversion | (0.95, 1.09) 0.639 | Excluded |
| independ | (0.91, 1.05) 0.587 | Excluded |
| **selfcontrol** | **(0.89, 1.02) 0.202** | **(0.90, 0.98) 0.007** |
| **anxiety** | **(0.89, 1.02) 0.158** | **(0.90, 1.00) 0.060** |
| novator | (0.95, 1.07) 0.766 | Excluded |

The model comparison confirms our findings and subsequent refining of the model to achieve the best fit. Variables that did not meet the assumption of proportional hazards were excluded as well as variables that were found to be non-significant.

## Predictions

We will now use our adjusted model to predict 3-year survival rates between two employees, one age 25 and one age 55. All other variables will remain the same.

```{r}
new_hire <- data.frame(age = c(25,55), industry = c("IT","IT"), stag= c(36, 36), event= c(0,0), profession = c("manage","manage"), traffic = c("empjs","empjs"), greywage = c("white","white"), way = c("car","car"), independ = c(5,5), selfcontrol = c(5,5), anxiety = c(5,5))

preds <- predict(model1, newdata = new_hire, type = "survival", se.fit = TRUE)

new_hire$prob <- preds$fit
new_hire$lcl <- preds$fit - 1.96*preds$se.fit
new_hire$ucl <- preds$fit + 1.96*preds$se.fit
new_hire
```
Our results show a survival probability of 38% for the 25-year old, and 17% percent for the 55-year old.

# Conclusion

# References

::: {#refs}
:::
